Physics inspired priors for neutron star mergers improves equation of state constraints

### 2025-02-20 paper.tex
minor cleanup.

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}

\documentclass[twocolumn]{aastex631}
%\usepackage{siunitx}
\usepackage{amsmath}

\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}

@@

	% \title{Microphysics and astrophysics inspired priors for neutron star mergers}
	\title{Physics inspired priors for neutron star mergers improves equation of state constraints}

@@

	\author{Christian Ecker}
	\affiliation{Institut f\"ur Theoretische Physik, Max-von-Laue-Strasse 1, 60438 Frankfurt, Germany}

	\author{Luciano Rezzolla}
	\affiliation{Institut f\"ur Theoretische Physik, Max-von-Laue-Strasse 1, 60438 Frankfurt, Germany}
	\affiliation{Frankfurt Institute for Advanced Studies, Ruth-Moufang-Strasse 1, 60438 Frankfurt, Germany}
	\affiliation{School of Mathematics, Trinity College, Dublin 2, Ireland}

@@

	\begin{abstract}
		Gravitational-wave astronomy shows great promise in determining nuclear physics in a regime not accessible to terrestrial experiments. 
		We introduce physics-inspired priors derived from a combination of constraints from nuclear-physics experiments, perturbative quantum chromodynamics calculations, and astrophysical constraints.
		When these priors are used in gravitational-wave astrophysical inference, we show significant improvement on nuclear equation of state constraints. 
		Applying these to the first two gravitational-wave binary neutron star mergers GW170817 and GW190425, constraints on the radius of a $1.4\,M_\odot$ neutron star improves from $R=\red{XX^{+YY}_{-ZZ}}\,{\rm km}$ to $\red{XX^{+YY}_{-ZZ}}\,{\rm km}$ (90\% confidence intervals).
		We also show these priors can be used to perform model selection between binary neutron star and neutron star-black hole mergers, although the results for GW190425 are incolnclusive with a Bayes factor of 1.33 in favour of the binary neutron star merger hypothesis, representing only marginal evidence.
		We advocate for these physics-inspired priors to be used as standard in the literature, and provide open-source code for reproducibility and further adaptation of the method.
	\end{abstract}

@@

	When performing gravitational-wave astrophysical inference of binary neutron star mergers~\citep[e.g.,][]{abbott17_170817observation,abbott18_170817EOS,abbott19_170817Properties,abbott20_190425}, it is common to use `agnostic' priors on their gravitational-wave observables, for example, uniform priors on chirp mass $\mathcal{M}$ and the dimensionless tidal deformability parameter $\tilde{\Lambda}$. However, \citet{altiparmak22} showed that the above conditions imply highly correlated probability distributions in $\mathcal{M}-\tilde{\Lambda}$ parameter space.

In this work, we advocate using these physics-inspired priors joint
priors on $\mathcal{M}$ and $\tilde{\Lambda}$, which are the two best
measured and most informative parameters related to the neutron star
equation of state. We show an improvement on the measurement of the
marginalized posterior on $\tilde{\Lambda}$ from the gravitational-wave
observation of GW170817 from $\tilde{\Lambda}=\red{XX}$ with the agnostic
prior to $\red{YY}$ with the new physics-inspired prior. This corresponds
to an improvement in the neutron star radius measurement of an equivalent
$[1.4]\,{M_\odot}$ star from $\red{R_{1.4}}= $ to $\red{\ldots}$ For
GW190425, the improvement is \red{\ldots}.

@@

	\red{Methodology for the speed of sound parameterisation, and eos generation and constraints should go here. }

	Our setup closely follows that of~~\citet{Altiparmak:2022}, which we briefly review here. 
	We begin by constructing a large set of EOSs by stitching together various components.
	At the lowest densities ($n<0.5 n_s$), we adopt the Baym-Pethick-Sutherland (BPS) model~\citep{Baym71} for the neutron-star crust.
	In the range $0.5 n_s < n < 1.1 n_s$, we randomly sample polytropes to span the range between the softest and stiffest EOSs from~\citet{Hebeler:2013nza}.
	At high densities ($n \gtrsim 40,n_s$), corresponding to a baryon chemical potential of $\mu = 2.6,{\rm GeV}$, we impose the perturbative QCD constraint from~\citet{Fraga2014} on the pressure $p(X, \mu)$ of cold quark matter, with the renormalization scale parameter $X$ sampled uniformly in $[1,4]$.

	For the intermediate density range ($1.1 n_s < n \lesssim 40 n_s$), we use the parametrization method of~\citet{Annala2019}, which models the sound speed as a function of the chemical potential, $c_s^2(\mu)$, using piecewise-linear segments:
	\begin{equation} \label{eq:cs2}
		c_s^2(\mu) = \frac{\left(\mu_{i+1}-\mu \right) c_{s,i}^2 + \left(\mu - \mu_i \right) c_{s,i+1}^2}{\mu_{i+1}-\mu_i}~, 
	\end{equation}
	where $\mu_i$ and $c_{s,i}^2$ are parameters defining the $i$-th segment in the range $\mu_i \leq \mu \leq \mu_{i+1}$.
	Throughout this work, we adopt natural units where $c=G=1$.

	The number density follows as 
	\begin{equation} \label{eq:n}
		n(\mu) = n_1 \exp \left({\int_{\mu_1}^\mu \frac{d\mu^\prime}{\mu^\prime c_s^2(\mu^\prime)}}\right)~, 
	\end{equation} where $n_1 = 1.1,n_s$, and $\mu_1 = \mu(n_1)$ is set by the corresponding polytropic EOS.
	The pressure is then obtained via
	\begin{equation} \label{eq:p}
		p(\mu) = p_1 + \int_{\mu_1}^\mu d\mu^\prime , n(\mu^\prime)~,
	\end{equation}
	where the integration constant $p_1$ matches the pressure of the polytrope at $n = n_1$.
	We numerically integrate Eq.~\eqref{eq:p} using seven segments for $c_s^2(\mu)$.

	Using this framework, we generate $XXX$ EOSs by randomly selecting the maximum sound speed $c_{s,{\rm max}}^2 \in [0,1]$ and uniformly sampling the free parameters $\mu_i \in [\mu_1, \mu_{N+1}]$ (where $\mu_{N+1} = 2.6\rm GeV$) and $c_{s,i}^2 \in [0, c_{s,{\rm max}}^2]$. These EOSs are consistent with nuclear theory and perturbative QCD uncertainties.

	To incorporate astrophysical constraints, we solve the Tolman-Oppenheimer-Volkoff (TOV) equations for each EOS and retain only those satisfying the mass measurements of J0348+0432~\citep{Antoniadis2013} ($M = 2.01\pm 0.04,M_{\odot}$) and J0740+6620~\citep{Cromartie2019, Fonseca2021} ($M = 2.08 \pm 0.07,M_{\odot}$), discarding EOSs with a maximum mass $M_{\rm TOV} < 2.0,M_{\odot}$. We also impose NICER radius constraints from J0740+6620~\citep{Miller2021, Riley2021} and J0030+0451~\citep{Riley2019, MCMiller2019b}, rejecting EOSs with $R < 10.75$ km at $M = 2.0,M_{\odot}$ or $R < 10.8$ km at $M = 1.1,M_{\odot}$.

	In the neutron star binary case, the binary tidal deformability is computed as
	\begin{equation} 
		\tilde{\Lambda}_{\rm NSNS} = \frac{16}{13} \frac{ (12M_2 + M_1) M_1^4 \Lambda_1 + (12M_1 + M_2) M_2^4 \Lambda_2}{(M_1 + M_2)^5}~,
	\end{equation}
	where $M_i$, $R_i$, and $\Lambda_i = \frac{2}{3} k_2 \left( R_i/M_i \right)^5$ ($i=1,2$) denote the component masses, radii, and tidal deformabilities, with $k_2$ being the second tidal Love number.
	The chirp mass is given by $\mathcal{M}_{\rm chirp} = (M_1 M_2)^{3/5} (M_1+M_2)^{-1/5}$ and we in the following we will use the mass ratio defined as $q = M_2/M_1$.

	In the mixed binary case, i.e., where one binary component is a neutron star and the other one a black hole, the formula for the binary tidal deformability simplifies, because black holes have zero tidal deformability $\Lambda_{\rm BH}=0$.
	The binary tidal deformability in this case becomes
	\begin{equation} 
		\tilde{\Lambda}_{\rm NSBH} = \frac{16}{13} \frac{ (12M_{\rm BH} + M_{\rm NS}) M_{\rm NS}^4 \Lambda_{\rm NS}}{(M_{\rm NS} + M_{\rm BH})^5}~,
	\end{equation}
	where $M_{\rm NS}$ and $\Lambda_{\rm NS}$ are the nuetron star mass and tidal defromability, respectively, and $M_{\rm BH}$ the black hole mass. The expressions for the chirp mass and mass ratio remain intact, with the understanding that $M_1=M_{\rm NS}$ and $M_2=M_{\rm BH}$.

>>>>

\documentclass[twocolumn]{aastex631}

\usepackage{amsmath}

\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\CE}[1]{\textcolor{violet}{CE: #1}}

@@

\title{Physics inspired priors for neutron star mergers improves equation of state constraints}

@@

\author[0000-0002-8669-4300]{Christian Ecker}
\affiliation{Institut f\"ur Theoretische Physik, Max-von-Laue-Strasse 1, 60438 Frankfurt, Germany}

\author[0000-0002-1330-7103]{Luciano Rezzolla}
\affiliation{Institut f\"ur Theoretische Physik, Max-von-Laue-Strasse 1, 60438 Frankfurt, Germany}
\affiliation{Frankfurt Institute for Advanced Studies, Ruth-Moufang-Strasse 1, 60438 Frankfurt, Germany}
\affiliation{School of Mathematics, Trinity College, Dublin 2, Ireland}

@@

\begin{abstract}
 Gravitational-wave astronomy shows great promise in determining nuclear physics in a regime not accessible to terrestrial experiments. 
 We introduce physics-inspired priors informed by nuclear theory and perturbative quantum chromodynamics calculations as well as astrophysical measurements of neutron star masses and radii.
 When these priors are used in gravitational-wave astrophysical inference, we show significant improvement on nuclear equation of state constraints. 
 Applying these to the first two gravitational-wave binary neutron star mergers GW170817 and GW190425, constraints on the radius of a $1.4\,M_\odot$ neutron star improves from $R=\red{XX^{+YY}_{-ZZ}}\,{\rm km}$ to $\red{XX^{+YY}_{-ZZ}}\,{\rm km}$ (90\% confidence intervals).
 We also show these priors can be used to perform model selection between binary neutron star and neutron star-black hole mergers, although the results for GW190425 are incolnclusive with a Bayes factor of $1.33$ in favour of the binary neutron star merger hypothesis, representing only marginal evidence.
 We advocate for these physics-inspired priors to be used as standard in the literature, and provide open-source code for reproducibility and further adaptation of the method.
 \CE{The abstract seems a bit optimisitc. Can we actually claim an improvement of the NS EOS and radius constraints?}
\end{abstract}

@@

When performing gravitational-wave astrophysical inference of binary neutron star mergers~\citep[e.g.,][]{abbott17_170817observation,abbott18_170817EOS,abbott19_170817Properties,abbott20_190425}, it is common to use `agnostic' priors on their gravitational-wave observables, for example, uniform priors on chirp mass $\mathcal{M}$ and the dimensionless tidal deformability parameter $\tilde{\Lambda}$. 
However, \citet{altiparmak22,Ecker:2022dlg} showed that the above conditions imply highly correlated probability distributions in $\mathcal{M}-\tilde{\Lambda}$ parameter space.

In this work, we advocate using these physics-inspired priors joint priors on $\mathcal{M}$ and $\tilde{\Lambda}$, which are the two best measured and most informative parameters related to the neutron star equation of state (EOS).
We show an improvement on the measurement of the marginalized posterior on $\tilde{\Lambda}$ from the gravitational-wave observation of GW170817 from $\tilde{\Lambda}=\red{XX}$ with the agnostic prior to $\red{YY}$ with the new physics-inspired prior.
This corresponds to an improvement in the neutron star radius measurement of an equivalent $[1.4]\,{M_\odot}$ star from $\red{R_{1.4}}= $ to $\red{\ldots}$ \CE{How to quantify this improvement?}.
For GW190425, the improvement is \red{\ldots}.

@@

Our setup closely follows that of~\citet{Altiparmak:2022}, which we briefly review here. 
We begin by constructing a large set of EOSs by stitching together various components.
At the lowest densities ($n<0.5\,n_s$), with $n_s=0.16{\rm fm}^-3$ the nuclear saturation density, we adopt the Baym-Pethick-Sutherland (BPS) model~\citep{Baym71} for the neutron-star crust.
In the range $0.5\,n_s < n < 1.1\,n_s$, we randomly sample polytropes to span the range between the softest and stiffest EOSs from~\citet{Hebeler:2013nza}.
At high densities ($n \gtrsim 40\,n_s$), corresponding to a baryon chemical potential of $\mu = 2.6\,\rm GeV$, we impose the perturbative QCD constraint from~\citet{Fraga2014} on the pressure $p(X, \mu)$ of cold quark matter, with the renormalization scale parameter $X$ sampled uniformly in $[1,4]$.

For the intermediate density range ($1.1\,n_s < n \lesssim 40\,n_s$), we use the parametrization method of~\citet{Annala2019}, which models the sound speed as a function of the chemical potential, $c_s^2(\mu)$, using piecewise-linear segments:
\begin{equation} \label{eq:cs2}
 c_s^2(\mu) = \frac{\left(\mu_{i+1}-\mu \right) c_{s,i}^2 + \left(\mu - \mu_i \right) c_{s,i+1}^2}{\mu_{i+1}-\mu_i}\,, 
\end{equation}
where $\mu_i$ and $c_{s,i}^2$ are parameters defining the $i$-th segment in the range $\mu_i \leq \mu \leq \mu_{i+1}$.
Throughout this work, we adopt natural units where $c=G=1$.

The number density follows as 
\begin{equation} \label{eq:n}
 n(\mu) = n_1 \exp \left({\int_{\mu_1}^\mu \frac{d\mu^\prime}{\mu^\prime c_s^2(\mu^\prime)}}\right)\,, 
\end{equation} 
where $n_1 = 1.1\,n_s$, and $\mu_1 = \mu(n_1)$ is set by the corresponding polytropic EOS.
The pressure is then obtained via
\begin{equation} \label{eq:p}
 p(\mu) = p_1 + \int_{\mu_1}^\mu d\mu^\prime , n(\mu^\prime)~,
\end{equation}
where the integration constant $p_1$ matches the pressure of the polytrope at $n = n_1$.
We numerically integrate Eq.~\eqref{eq:p} using seven segments for $c_s^2(\mu)$.

Using this framework, we generate $\approx 10^6$ EOSs by randomly selecting the maximum sound speed $c_{s,{\rm max}}^2 \in [0,1]$ and uniformly sampling the free parameters $\mu_i \in [\mu_1, \mu_{N+1}]$ (where $\mu_{N+1} = 2.6\,\rm GeV$) and $c_{s,i}^2 \in [0, c_{s,{\rm max}}^2]$.
These EOSs are by construction consistent with nuclear theory and perturbative QCD uncertainties.

To incorporate astrophysical constraints, we solve the Tolman-Oppenheimer-Volkoff (TOV) equations for each EOS and retain only those satisfying the mass measurements of J0348+0432~\citep{Antoniadis2013} ($M = 2.01\pm 0.04\,M_{\odot}$) and J0740+6620~\citep{Cromartie2019, Fonseca2021} ($M = 2.08 \pm 0.07\,M_{\odot}$), discarding EOSs with a maximum mass $M_{\rm TOV} < 2.0\,M_{\odot}$. 
In addition, we impose NICER radius constraints from J0740+6620~\citep{Miller2021, Riley2021} and J0030+0451~\citep{Riley2019, MCMiller2019b}, rejecting EOSs with $R < 10.75\,\rm km$ at $M = 2.0\,M_{\odot}$ or $R < 10.8\,\rm km$ at $M = 1.1\,M_{\odot}$.

In the neutron star binary case, the binary tidal deformability is computed as
\begin{equation} 
 \tilde{\Lambda}_{\rm NSNS} = \frac{16}{13} \frac{ (12M_2 + M_1) M_1^4 \Lambda_1 + (12M_1 + M_2) M_2^4 \Lambda_2}{(M_1 + M_2)^5}\,,
\end{equation}
where $M_i$, $R_i$, and $\Lambda_i = \frac{2}{3} k_2 \left( R_i/M_i \right)^5$ ($i=1,2$) denote the component masses, radii, and tidal deformabilities, with $k_2$ being the second tidal Love number.
The chirp mass is given by $\mathcal{M}_{\rm chirp} = (M_1 M_2)^{3/5} (M_1+M_2)^{-1/5}$ and we in the following we will use the mass ratio defined as $q = M_2/M_1$.

In the mixed binary case, i.e., where one binary component is a neutron star and the other one a black hole, the formula for the binary tidal deformability simplifies, because black holes have zero tidal deformability $\Lambda_{\rm BH}=0$.
The binary tidal deformability in this case becomes
\begin{equation} 
 \tilde{\Lambda}_{\rm NSBH} = \frac{16}{13} \frac{ (12M_{\rm BH} + M_{\rm NS}) M_{\rm NS}^4 \Lambda_{\rm NS}}{(M_{\rm NS} + M_{\rm BH})^5}\,,
 \end{equation}
where $M_{\rm NS}$ and $\Lambda_{\rm NS}$ are the nuetron star mass and tidal defromability, respectively, and $M_{\rm BH}$ the black hole mass. 
The expressions for the chirp mass and mass ratio remain intact, with the understanding that $M_1=M_{\rm NS}$ and $M_2=M_{\rm BH}$. 
Unlike in~\citet{altiparmak22,Ecker:2022dlg} we do not impose any GW informed constraints on the binary tidal deformability in the construction of our proior.
